Towards Systematic Specification and Verification of Fault-Tolerant Protocols

### 2016-07-05 doc/om-paper/paper.tex

There is substantial literature on verifying fault-tolerant distributed systems~\cite{}. Some of this work uses mechanical theorem-proving~\cite{}. Two problems with mechanical verification are that it requires specialized expertise and it is labor intensive. In contrast, model-checking is automated but has limited scalability. To address scalability, abstractions and optimizations are employed, both of which lead to problems. Verification abstractions are often ad-hoc and system-specific. For example, to model a node that can broadcast the values $\{0, 1\}$, or no value if when faulty, one might define an enumerated type of broadcast values sent such as

@@

However, the enumeration conflates environment artifacts (i.e., faults) and system artifacts. Conflating them makes formal model-to-code correspondence more difficult to achieve. An example of a verification optimization is to model message passing as shared state between nodes. While these optimizations might be safe under particular assumptions, they can lead to unsound models when a system or its environment is modified.

@@

We demonstrate the model with an extended case-study, verifying the Oral Messages (1) protocol (OM(1))~\cite{}. While our model is more detailed---it explicitly models channels---and also more general---it allows bounded asynchrony between nodes---it is the most scalable model-checking verification of the protocol.

@@

\subsection{Calendar Automata}\label{sec:calendar}
\emph{Timed automata} are specialized formalisms for verifying real-time systems~\cite{}. Time automata assume the existence of continuously-varying real-valued variables, known as \emph{real-time clocks} in a solver. While successful timed automata model-checkers exist, they are specialized for verifying real-time systems in which the only infinite-valued variables are the real-time clocks. Modern general-purpose model-checkers like \lee{name some} use SMT solvers for general infinite-state system verification that is not limited to real-time clocks.

@@

Let $c$ be an event calendar and $c_i, c_j \in c$ be calendar events. Define an ordering on event calendars such that $c_i \leq c_j$ iff $t_i \leq t_j$, and $min(c) = \{ c_i | \forall c_j \in c, \, c_i \leq c_j  \}$ are the minimum elements of $c$.

Let a transition system $\mathcal{M} = (S, I, \rightarrow)$, be a set of states $S$, a set of initial states $I \subseteq S$, and a transition relation $\rightarrow \subseteq S \times S$. We implicitly assume a set of state variables such that a state $\sigma \in S$ is a total function that maps state variables to values. We distinguish two special state variables in a transition system: (1) $now \in \mathbb{R}$ denotes the current time in the state, and (2) $c$ is an event calendar.

@@

\begin{enumerate}
\item Time is initialized to be less than or equal to every calendar timeout: $\forall \sigma \in I, \, \forall (e_i, t_i) \in \sigma(c)$, $\sigma(now) \leq t_i$.
\item In all states, every calendar event occurs no sooner than the current time: $\forall \sigma \in S \, \forall (e_i, t_i) \in \sigma(c)$, $t_i \geq \sigma(now)$.
\item In all states, if the current time is strictly less than every calendar event, then the only enabled transition is a \emph{time progress} update: $\forall \sigma \in S, \, \forall (e_i, t_i) \in \sigma(c)$, if $\sigma(now) < t_i$, then $\forall \sigma'$ such that $\sigma \rightarrow \sigma'$, $\sigma' = \sigma[now := min(c)]$.
\item In all states, if the current time equals a timeout, then the only transition enabled is a calendar event update associated with the timeout: $\forall \sigma \in S, \, \exists (e_i, t_i) \in \sigma(c)$ such that $\sigma(now) = t_i$, then $\forall \sigma'$ such that $\sigma \rightarrow \sigma'$, $\sigma'(now) = \sigma(now)$, $\sigma'(c_j) = \sigma(c_j)$ for all $c_j \in \sigma(c)$ such that $c_j \neq c_i$, and $\sigma'(t_i) > \sigma(t_i)$.
\end{enumerate}

@@

Assume processes are indexed with the natural numbers $\mathbb{N}$. A \emph{channel} from process $i$ to $j$ is an ordered pair $(i,j)$. Define a set of messages $Msg$, with a designated element $null \in Msg$ denoting the absence of a message. Given a channel and a timeout, let $send$ return the message sent on the channel at that time:
$$send : \mathbb{N} \times \mathbb{N} -> \mathbb{R} -> Msg$$
So $send(i, j, t)$ returns the message that $i$ sends to $j$ at time $t$. Likewise, let
$$recv : \mathbb{N} \times \mathbb{N} -> \mathbb{R} -> Msg$$
compute the message received on that channel at that time, so that $recv(i, j, t)$ returns the message received by $j$ from $i$ at time $t$.

We require that messages received were previously sent and not previously received: if $recv(i, j, t) \neq null$, then $\exists t'$ such that $send(i, j, t')$ where $t' < t$, $recv(i, j, t) = send(i, j, t')$, and $\neg\exists t''$ such that $t' < t'' < t$ and $recv(i, j, t'') = recv(i, j, t)$

@@

% ------------------------------------------------------------
\section{The Synchronous Kibitzer}
The typical approach to injecting faults into a formal model of a distributed system is to add a new state variables to each process representing its fault state. Then additional state variables are added to represent the nondeterminism introduced by faults into the state or messages by a process. Finally, additional transitions are added to each process to modify the process's state or messages based on it's current fault state.

@@

To address these problems, we introduce a synchronous observer that injects faults that we call a \emph{synchronous kibitzer}. The kibitzer decomposes the state and transitions associated with the fault model from the system.

For the sake of concreteness, we introduce
 under the hybrid fault model of Thambidurai and Park~\cite{}. The fault model distinguishes Byzantine (or arbitrary), symmetric, and manifest faults. This fault model applies to broadcast systems, in which a process is expected to broad the same value to multiple receivers. A Byzantine faulty

% ------------------------------------------------------------
\section{Abstract State Machines}\label{sec:abstract}

>>>>

There is substantial literature on verifying fault-tolerant distributed systems~\cite{}. Some of this work uses mechanical theorem-proving~\cite{}. Two problems with mechanical verification are that it requires specialized expertise and it is labor intensive. In contrast, model-checking is automated but has limited scalability. To address scalability, abstractions and optimizations are employed, both of which lead to problems. Verification abstractions are often ad-hoc and system-specific. For example, to model a process that can broadcast the values $\{0, 1\}$, or no value if when faulty, one might define an enumerated type of broadcast values sent such as

@@

However, the enumeration conflates environment artifacts (i.e., faults) and system artifacts. Conflating them makes formal model-to-code correspondence more difficult to achieve. An example of a verification optimization is to model message passing as shared state between processes. While these optimizations might be safe under particular assumptions, they can lead to unsound models when a system or its environment is modified.

@@

We demonstrate the model with an extended case-study, verifying the Oral Messages (1) protocol (OM(1))~\cite{}. While our model is more detailed---it explicitly models channels---and also more general---it allows bounded asynchrony between processes---it is the most scalable model-checking verification of the protocol.

@@

\subsection{Calendar Automata}\label{sec:calendar}\label{sec:calendar}
\emph{Timed automata} are specialized formalisms for verifying real-time systems~\cite{}. Time automata assume the existence of continuously-varying real-valued variables, known as \emph{real-time clocks} in a solver. While successful timed automata model-checkers exist, they are specialized for verifying real-time systems in which the only infinite-valued variables are the real-time clocks. Modern general-purpose model-checkers like \lee{name some} use SMT solvers for general infinite-state system verification that is not limited to real-time clocks.

@@

Let $cal$ be an event calendar and $c_i, c_j \in c$ be calendar events. Define an ordering on event calendars such that $c_i \leq c_j$ iff $t_i \leq t_j$, and $min(c) = \{ c_i | \forall c_j \in cal, \, c_i \leq c_j  \}$ are the minimum elements of $cal$.

Let a transition system $\mathcal{M} = (S, I, \rightarrow)$, be a set of states $S$, a set of initial states $I \subseteq S$, and a transition relation $\rightarrow \subseteq S \times S$. We implicitly assume a set of state variables such that a state $\sigma \in S$ is a total function that maps state variables to values. We distinguish two special state variables in a transition system: (1) $now \in \mathbb{R}$ denotes the current time in the state, and (2) $cal$ is an event calendar.

@@

\begin{enumerate}
\item Time is initialized to be less than or equal to every calendar timeout: $\forall \sigma \in I, \, \forall (e_i, t_i) \in \sigma(cal)$, $\sigma(now) \leq t_i$.

\item In all states, every calendar event occurs no sooner than the current time: $\forall \sigma \in S \, \forall (e_i, t_i) \in \sigma(cal)$, $t_i \geq \sigma(now)$.

\item In all states, if the current time is strictly less than every calendar event, then the only enabled transition is a \emph{time progress} update: $\forall \sigma \in S, \, \forall (e_i, t_i) \in \sigma(cal)$, if $\sigma(now) < t_i$, then $\forall \sigma'$ such that $\sigma \rightarrow \sigma'$, $\sigma' = \sigma[now := min(c)]$.

\item In all states, if the current time equals a timeout, then the only transition enabled is a calendar event update associated with the timeout: $\forall \sigma \in S, \, \exists (e_i, t_i) \in \sigma(cal)$ such that $\sigma(now) = t_i$, then $\forall \sigma'$ such that $\sigma \rightarrow \sigma'$, $\sigma'(now) = \sigma(now)$, $\sigma'(c_j) = \sigma(c_j)$ for all $c_j \in \sigma(c)$ such that $c_j \neq c_i$, and $\sigma'(t_i) > \sigma(t_i)$.
\end{enumerate}

@@

Assume processes are indexed from a finite set $Id \subset \mathbb{N}$. A \emph{channel} from process $i$ to $j$ is an ordered pair $(i,j)$. Define a set of messages $Msg$, with a designated element $null \in Msg$ denoting the absence of a message. Given a channel and a timeout, let $send$ return the message sent on the channel at that time:
$$send : Id \times Id \rightarrow \mathbb{R} \rightarrow Msg$$
So $send(i, j, t)$ returns the message that $i$ sends to $j$ at time $t$. Likewise, let
$$recv : Id \times Id \rightarrow \mathbb{R} \rightarrow Msg$$
compute the message received on that channel at that time, so that $recv(i, j, t)$ returns the message received by $j$ from $i$ at time $t$.

In the non-faulty case, we require that messages received were previously sent and not previously received: if $recv(i, j, t) \neq null$, then $\exists t'$ such that $send(i, j, t')$ where $t' < t$, $recv(i, j, t) = send(i, j, t')$, and $\neg\exists t''$ such that $t' < t'' < t$ and $recv(i, j, t'') = recv(i, j, t)$. (We address faults in Section~\ref{sec:kibitzer}.)

@@

% ------------------------------------------------------------
\section{The Synchronous Kibitzer}\label{sec:kibitzer}
The typical approach to injecting faults into a formal model of a distributed system is to add a new state variables to each process representing its fault state. Then additional state variables are added to represent the nondeterminism introduced by faults into the state or messages by a process. Finally, additional transitions are added to each process to modify the process's state or messages based on it's current fault state.

@@

To address these problems, we introduce a synchronous observer that injects faults that we call a \emph{synchronous kibitzer}. The kibitzer decomposes the state and transitions associated with the fault model from the system. For the sake of concreteness, we introduce a particular fault model, the hybrid fault model of Thambidurai and Park~\cite{}. The fault model distinguishes Byzantine (or arbitrary), symmetric, and manifest faults. This fault model applies to broadcast systems, in which a process is expected to broad the same value to multiple receivers. A \emph{Byzantine fault} is one in which a process that is intended to broadcast the same value to other processes may instead broadcast arbitrary values to different receivers (including no value or the correct value). A \emph{symmetric fault} is one in which a process may broadcast the same, but incorrect, value to other processes, and a \emph{manifest fault} is one in which a process's broadcast fault is detectable by the receivers; e.g., by performing a cyclic redundancy check (CRC) or because the value arrives outside of a predetermined window.

In specifying a fault-tolerant system, we must also specify a \emph{maximum fault assumption} (MFA), which specifies the maximum number of of each kind of fault that may be present in the system for some system property to hold. Define a enumeration of fault types $$\textnormal{FAULT\_TYPE} = \{none, byz, sym, man\}$$. As in the previous section, let $Id \subset \mathbb{N}$ be a finite set of indices, and let the variable $$faults: Id \rightarrow \textnormal{FAULT\_TYPE}$$ range over possible mappings from process indices to faults.

The hybrid fault model assumes a broadcast model of communication.
% Define $rnd: \mathbb{R} \rightarrow \mathbb{N}$ such that if $rnd(t_0) < rnd(t_1)$, then $t_0 < t_1$.
A $broadcast: Id \rightarrow 2^{Id} \rightarrow \mathbb{R} \rightarrow Msg \rightarrow 2^{E}$ takes a sender, a set of receivers, a real-time, and a message to send each receiver, and returns a set of calendar events:
$$broadcast(i, R, t, m) = \{m | j \in R \textnormal{ and } send(i, j, t) = m\}$$

\lee{msg in SAL is recv here}

With this machinery, we can define the semantics of faults by constraining the relationship between a message broadcast and the values received by the recipients. For a nonfaulty process that broadcasts, every recipient receives the sent message.
\begin{align*}
&nonfaulty\_constraint =\\
  &\quad \forall i, j \in Id, t \in \mathbb{R}.\\
  &\quad\quad (faults(i) = none\\
  &\quad\quad\quad \textnormal{ implies } recv(i, j, t) = send(i, j, t)
\end{align*}
\noindent
For symmetric faults, we there is no requirement that the messages sent are the ones received, only that every recipient receives the same value.
\begin{align*}
&sym\_constraint =\\
  &\quad \forall i, j, k \in Id, t \in \mathbb{R}.\\
  &\quad\quad (faults(i) = sym \textnormal{ and } broadcast(i, \{j, k\}, t, m))\\
  &\quad\quad\quad \textnormal{ implies } recv(i, j, t) = recv(i, k, t)
\end{align*}
\noindent
Byzantine faults are left completely unconstrained.

% ------------------------------------------------------------
\section{Abstract State Machines}\label{sec:abstract}\label{sec:asms}
When model-checking a system, counterexamples to properties are traces of states. A state is a collection of variable values. For industrial-scale models, there are many 10s or even 100s of state variables. So a single state includes assignments to each of the variables, and counterexample trace may be many 10s of states in length. In practice, counterexamples depend on a small number of state variables. The user has to manually separate the state variables that matter from those that do not for a specific property. It is not enough prevent variables that stutter from displaying since some state variables might be updated without affecting the property. The problem is particularly pertinent when model-checking models that have sufficient detail for implementation synthesis, as is the goal in our work.

Systems are usually designed in modes such predicates on state variables are parameterized on which mode the system is in. The transition between modes implements a finite state-machine. The modes and the state-machine are usually implicit, but a property violation usually requires an unexpected mode transition. Abstracting counterexample traces over all of the state variables by traces over the modes helps to identify why a property fails quickly.

Our solution to the problem of constructing a mode-based state machine uses abstract state machines (ASMs). ASMs are a generalization of finite state machines originally developed by Gurevich to generalize the Church-Turing thesis for abstract data structures~\cite{}. Since then, ASMs have become an engineering technique for simulating a system at different levels of abstraction.

\lee{http://www2.informatik.hu-berlin.de/sam/preprint/reisig203.pdf}
More precisely, we implement a sequential small-step ASM~\cite{}.
\begin{definition}[Sequential small-step ASM]
Let $V$ be a finite set of state variables. Let $s$ be an assignment of variables to values, and let $b$ be a relation Å
\end{definition}
